Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2019-10-29 paper.Rnw
appendix plus

\documentclass{article}
\usepackage[stable]{footmisc}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[OT1]{fontenc}
\usepackage{amsthm,amsmath,amsfonts}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{xr}
\usepackage{tikz}
\usepackage{authblk}

%\usepackage{endfloat}

@@

@@

<<init,include=FALSE>>=
library(knitr)
opts_chunk$set(
echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
)
pn <- function(...) prettyNum(...,big.mark=',')
library(xtable)
@

@@

<<loadStuff,cache=FALSE>>=
load('output/analysis.RData')
@

@@

Our intention is to develop, demonstrate, and contrast these three
approaches to modeling log data.
We hope that this paper can serve as a starting point for
applied educational researchers interested in mining this rich
resource, and that this paper helps spur methodological interest in
this problem.

@@

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{h_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately \Sexpr{round(mixProp*100)}\% of students as
high-hint users, so we chose $c$ as the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$,
approximately \Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
This definition agreed with the model's classification in
approximately \Sexpr{round(agree*100)}\% of cases.

@@

Let $Y_i$ represent subject $i$'s posttest score, and let $Z_i$ represent
$i$'s treatment status (i.e. the CTA1 or control group).
Following \citet{neyman} and \citet{rubin}, let $Y_i(Z=1)$ be
the posttest score student $i$ would achieve were $i$ (perhaps
counterfactually) assigned to the treatment condition, and
$Y_i(Z=0)$ the score were $i$ assigned to control.
Then the CTA1 treatment effect for student $i$ is $Y_i(Z=1)-Y_i(Z=0)$.
A similar structure applies to hint request for students in the
treatment group.
For these students, let the potential outcomes be $Y_i(Z=1,H=1)$ and
$Y_i(Z=1,H=0)$ or $Y_i(H=1)$ and $Y_i(H=0)$ for short.
For a study measuring hint requests as a continuous variable, we may
define $Y_i(\hcont)$, the posttest score $i$ would exhibit were $i$ to
request hints with proclivity $\hcont$.
This structure implicitly assumes ``non-interference,'' that one
student's treatment assignment $Z$ or hint proclivity $H$ does not
affect another student's posttest scores.

@@

\subsection{The Observational Study Approach: Interpretation and Discussion}
The observational study approach to log data seeks to estimate the
effect of a usage parameter, estimated from  computer log data, on
outcomes.
This approach does not exploit randomization---in fact, it discards
the control group entirely.
However, it depends heavily on another feature of RCTs: namely, an
outcome measured for all participants, in this case a posttest.
Causal identification derives not from randomization, but from an
untestable ignorability assumption, coupled with a sensitivity analysis.
In our example, we operationalized strong ignorability with a
propensity-score matching design, but other observational study
designs may apply as well.

We found that hint requests have a negative effect: students who
requested hints more often than the \Sexpr{round((1-mixProp)*100)}th
percentile score lower on the posttest, after controlling for observed covariates.
When hint usage is expressed numerically, we find a monotonically
decreasing relationship between hint requests and posttest scores---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
If, say, weaker students ask for hints more often and score lower
on the posttest regardless of hint usage, posttest scores will be
anticorrelated with hint requests.
This could even be the case after controlling for pretest scores and other
covariates, if these failed to capture all of the relevant dimensions
of student ability.
We found that an unobserved confounder that predicts both hint usage
and posttest scores just as strongly as the pretest could explain the
entire observed relationship.
The data, coupled with the existence of such a missing covariate, are
consistent with both positive and negative effects of hint requests.
On the other hand, covariates that are as important as pretests are
quite rare in practice---this confounding scenario might not be
plausible after all.
A less extreme missing confounder, predicting $Y$ and $H$ with half
the strength of the pretest, cannot explain the observed relationship,
and is only consistent with negative effects of hint requests.

Despite being hobbled by the lack of randomization, and the resultant
threat of confounding, the observational study approach provides weak
evidence that hints are harmful to posttest scores---a potentially
important finding.
This result, if confirmed in other scenarios and other analyses, may
inform student behavior in the tutor and future tutor designs.
However, since it ignores the control group from the CTA1 RCT, it does
not shed light on the CTA1 causal mechanisms---it does not explain the
CTA1 effect.
However, its design and results figure prominently in mediation
analysis, which does.

@@

Our solution was to dichotomize hint requests, as
described above, essentially re-defining the causal question.
\Sexpr{round((mixProp)*100)}\% of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the
\Sexpr{round((1-mixProp)*100)}th percentile of the treatment group.
To make sense of the respective roles of $H$ and $\hcont$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\hcont$.
In the control group all students are assigned $H=0$ and $\hcont=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\hcont$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\hcont=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

@@

\subsection{Mediation Analysis: Interpretation and Discussion}
Here we have used mediation analysis as a formal framework to
interpret the results of the observational study in Section
\ref{sec:observational} in terms of the overall CTA1 treatment effect.
Doing so requires some conceptual and statistical acrobatics, such as
relying on a dichotomized measurement of hint requests and strictly
counterfactual potential outcomes, but can nevertheless yield
interesting and interpretable results.
Specifically, our analysis showed that CTA1 affected students'
posttest scores in (at least) two opposite ways: by allowing hints, it
lowered their posttest scores, but its other mechanisms increased
their scores by an even greater amount.

Like the observational study, these results are subject to
confounding; indeed, the sensitivity analysis in the Section
\ref{sec:observational} suggests that these results, too, may be
explained due to a missing confounder.

\section{Principal Stratification}\label{sec:principalStratification}
The observational study approach and mediation analysis both rely on
an ignorability assumption to estimate the effect of $H$ on $Y$, in
the latter case framing this effect as one component of the overall
treatment effect.
Principal stratification \citep[PS;][]{frangakis} takes a different approach entirely.
In the PS approach, $Y(0)=Y(Z=0)$ and $Y(1)=Y(Z=1)$ are the only potential
outcomes for $Y$.
As in mediation analysis, hint behavior has potential outcomes as
well; modeling hint requests continuously with the $\hcont$ parameter
from (\ref{eq:rasch}) gives potential values $\hcont(1)$ and $\hcont(0)$,
though only $\hcont(1)$ is relevant in the CTA1 study.

The goal of principal stratification is to estimate ``principal
effects'':
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\hcont(1)=r]
\end{equation*}
This is the treatment effect for subjects who \emph{would} request
hints at level $\hcont(1)=r$, if assigned to the treatment condition.
Whereas the focus of the observational study and mediation
analysis was the relationship between $\hcont$ and posttest scores, the
focus of PS is the relationship between $\hcont$ and treatment effects.
Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\hcont(1)=r]$,
despite the fact that $\hcont(1)$ is never observed at the same time as
$Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
strictly counterfactual quantities, as in mediation analysis.
$\hcont(1)$ is defined for subjects in both randomized groups, but only
observed in the treatment group.

In previous sections $\hcont$ was dichotomized into $H$; such
dichotomization is not necessary here.
However, $\hcont$ has some disadvantages as a continuous measure of
hint usage.
For one, since it is essentially a sample mean over a subset of each student's
worked problems, much of its variance is driven by the total number of
problems students worked---in particular, its extreme values belong to
those students who barely used the tutor at all.
Further, it does not account for varying difficulty of the tutor's
problems.
As an alternative, we fit a
modified Rasch model \citep[e.g.][]{rasch},
\begin{equation}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems in each section are equally
likely to elicit a hint.

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the larger principal stratification model, replacing $\eta$ for
$\hcont$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

We modeled students' posttest scores $Y$ as
conditionally normal:
\begin{multline}\label{eq:outcomeSubmodel}
 Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim \\ \mathcal{N}\left(
\beta^Y_{0b[i]}+\bm{x}_i^T\bm{\beta}^Y+a\eta(1)+Z_i\tau(\eta(1))+\epsilon^{Yt}_{t[i]}+\epsilon^{Ys}_{s[i]},\sigma^Y_{Z[i]}\right)
\end{multline}
where $\beta^Y_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\beta}^Y$ are the
covariate coefficients, and $\epsilon^{Yt}$, and
$\epsilon^{Ys}$ are normally-distributed teacher and school random
intercepts.
The residual variance $\sigma^Y$ varies with treatment assignment $Z$;
this captures measurement error in $Y$, treatment effect heterogeneity
that is not linearly related to
$\eta(1)$, and other between-student variation in $Y$ that is not predicted by
the mean model.

@@

\subsection{Principal Stratification: Interpretation and Discussion}
Like in the observational study and in mediation analysis, the
principal stratification analysis found a negative correlation
between hint requesting and posttest scores.
Unlike those two methods, PS understands $\eta(1)$ as a
student-specific baseline characteristic---albeit one that only
manifests in the treatment group.
Students who request more hints need more help, and thus score lower
on the posttest.
On the other hand, the PS analysis suggests that it's these students
who also receive more help---perhaps due to CTA1 hints, or other
features of the curriculum---and hence benefit more from CTA1.

@@

On the other hand, the observational and mediational results may be
due not to causation but to confounding.
In contrast, PS does not depend on an ignorability assumption beyond
what is guaranteed by randomization.
Perhaps by recognizing $\eta(1)$ as hopelessly confounded and focusing
attention elsewhere, PS is able to arrive at the more reliable
conclusion that---whatever the mechanism may be---students who ask
more hints benefit more from the software.

Though PS does not depend on ignorability, it does depend on a highly
intricate and parametric modeling structure.
If any part of the model is sufficiently misspecified, the results may
be wrong.
Perhaps worse, \citet{feller2016principal} argued that even well-specified
models can give severely biased estimates.
Extensive model interrogation and checking is necessary for PS
analysis, but the complexity of the PS model makes this process
particularly difficult.
\citet{aoas} gives some examples of potentially fruitful model
checking procedures.

In contrast, the matching estimator in Section \ref{sec:observational}
is highly transparent.
Whatever intricacies went into formulating and fitting a propensity
score model or matching procedure are inconsequential if they produce
a convincing match.
The validity of the matching estimate rests entirely on the quality of
the match, which can be examined by checking covariate balance and
other relatively simple metrics.
The uncertainty in matching estimators due to possible confounding
might be offset by the simplicity in modeling assumptions.

\section{Which Strategy is Best?}

\begin{table}
\begin{tabular}{p{1in}p{2in}p{2in}}
& Observational Study/Mediation Analysis &Principal Stratification\\
\hline
Research question & Does $H$ help $Z$ affect $Y$? & Does $H$ help predict
                                           treatment effect of $Z$ on $Y$?\\
\hline
Role of $H$ & Causal & Predictive\\
\hline
Untestable assumption & $H$ is unconfounded & Distribution of $Y$
                                              given $H_T$ in control
                                              group\\
\hline
Other difficulties & Discrete $H$ necessary for mediation (in our
                     framework) & Model fitting can be opaque and
                                  erratic\\
\end{tabular}
\end{table}

@@

A second challenge is extrapolation in mediation analysis.
In RCTs comparing computerized educational technology with traditional
pedagogy, members of the control group do not have computer log data
or access to the mechanisms that the log data measures.
This forms a serious barrier to mediation analysis.
However, by dichotomizing the mediator and matching, we were able to
re-define the causal question in such a way as to make it (possibly) answerable.

@@

>>>>

\documentclass{article}
\usepackage[stable]{footmisc}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[OT1]{fontenc}
\usepackage{amsthm,amsmath,amsfonts}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{xr}
\usepackage{tikz}
\usepackage{authblk}
\usepackage[section]{placeins}
%\usepackage{endfloat}

@@

\renewenvironment{knitrout}{\begin{singlespace}}{\end{singlespace}}

@@

<<init,include=FALSE>>=
library(knitr)
opts_chunk$set(
echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
)
pn <- function(...) prettyNum(...,big.mark=',')
library(xtable)
library(english)

@

@@

<<loadStuff,cache=TRUE>>=
load('output/analysis.RData')
@

@@

Our intention is to develop, demonstrate, and contrast these three
approaches to modeling log data.
A rich, and often polemic, literature surrounds principal
stratification and mediation analysis
\citep[e.g.][]{rubin2004direct,vanderweele2011principal,pearl2011principal,mealli2012refreshing,vanderweele2012comments}.
\citet{vanderweele2008simple}, \citet{jo2008causal} and others have
derived formal equivalencies between parameters under certain
conditions.
Our contribution will build on this literature in two ways: first, by
providing a detailed demonstration of matching, mediation analysis,
and principal stratification on the same dataset, and second by
focusing on the applicability of these three methods in the specific
context of studying implementation data from an EdTech randomized
experiment.

@@

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{\bar{h}_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately \Sexpr{round(mixProp*100)}\% of students as
high-hint users, so we chose $c$ as the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$,
approximately \Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
This definition agreed with the model's classification in
approximately \Sexpr{round(agree*100)}\% of cases.

@@

Let $Y_i$ represent subject $i$'s posttest score, and let $Z_i$ represent
$i$'s treatment status (i.e. the CTA1 or control group).
Following \citet{neyman} and \citet{rubin}, let $Y_i(Z=1)$ be
the posttest score student $i$ would achieve were $i$ (perhaps
counterfactually) assigned to the treatment condition, and
$Y_i(Z=0)$ the score were $i$ assigned to control.
Then the CTA1 treatment effect for student $i$ is $Y_i(Z=1)-Y_i(Z=0)$.
A similar structure applies to hint request for students in the
treatment group.
For these students, let the potential outcomes be $Y_i(Z=1,H=1)$ and
$Y_i(Z=1,H=0)$ or $Y_i(H=1)$ and $Y_i(H=0)$ for short.
% For a study measuring hint requests as a continuous variable, we may
% define $Y_i(\bar{h})$, the posttest score $i$ would exhibit were $i$ to
% request hints with proclivity $\bar{h}$.
This structure implicitly assumes ``non-interference,'' that one
student's treatment assignment $Z$ or hint proclivity $H$ does not
affect another student's posttest scores.

@@

@@

Our solution was to dichotomize hint requests, as
described above, essentially re-defining the causal question.
\Sexpr{round((mixProp)*100)}\% of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the
\Sexpr{round((1-mixProp)*100)}th percentile of the treatment group.
To make sense of the respective roles of $H$ and $\bar{h}$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\bar{h}$.
In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\bar{h}$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

@@

\section{Principal Stratification}\label{sec:principalStratification}
The observational study approach and mediation analysis both rely on
an ignorability assumption to estimate the effect of $H$ on $Y$, in
the latter case framing this effect as one component of the overall
treatment effect.
Principal stratification \citep[PS;][]{frangakis} takes a different approach entirely.
In the PS approach, $Y(0)=Y(Z=0)$ and $Y(1)=Y(Z=1)$ are the only potential
outcomes for $Y$.
As in mediation analysis, hint behavior has potential outcomes as
well.
Specifically, the rate at which as student requests hints, $\bar{h}$,
has potential values $\bar{h}(1)$ and $\bar{h}(0)$,
though only $\bar{h}(1)$ is relevant in the CTA1 study.

The goal of principal stratification is to estimate ``principal
effects'':
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\bar{h}(1)=r]
\end{equation*}
This is the treatment effect for subjects who \emph{would} request
hints at level $\bar{h}(1)=r$, if assigned to the treatment condition.
Whereas the focus of the observational study and mediation
analysis was the relationship between $\bar{h}$ and posttest scores, the
focus of PS is the relationship between $\bar{h}$ and treatment effects.
Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\bar{h}(1)=r]$,
despite the fact that $\bar{h}(1)$ is never observed at the same time as
$Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
strictly counterfactual quantities, as in mediation analysis.
$\bar{h}(1)$ is defined for subjects in both randomized groups, but only
observed in the treatment group.

In previous sections $\bar{h}$ was dichotomized into $H$; such
dichotomization is not necessary here.
However, $\bar{h}$ has some disadvantages as a continuous measure of
hint usage.
For one, since it is essentially a sample mean over a subset of each student's
worked problems, much of its variance is driven by the total number of
problems students worked---in particular, its extreme values belong to
those students who barely used the tutor at all.
Further, it does not account for varying difficulty of the tutor's
problems.
As an alternative, we fit a
modified Rasch model \citep[e.g.][]{rasch},
\begin{equation}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems in each section are equally
likely to elicit a hint.

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the larger principal stratification model, replacing $\eta$ for
$\bar{h}$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

We modeled students' posttest scores $Y$ as
conditionally normal:
\begin{equation}\label{eq:outcomeSubmodel}
 Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim  \mathcal{N}\left(
\beta^Y_{0b[i]}+\bm{x}_i^T\bm{\beta}^Y+a\eta(1)+Z_i\tau(\eta(1))+\epsilon^{Yt}_{t[i]}+\epsilon^{Ys}_{s[i]},\sigma^Y_{Z[i]}\right)
\end{equation}
where $\beta^Y_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\beta}^Y$ are the
covariate coefficients, and $\epsilon^{Yt}$, and
$\epsilon^{Ys}$ are normally-distributed teacher and school random
intercepts.
The residual variance $\sigma^Y$ varies with treatment assignment $Z$;
this captures measurement error in $Y$, treatment effect heterogeneity
that is not linearly related to
$\eta(1)$, and other between-student variation in $Y$ that is not predicted by
the mean model.

@@

\section{Comparing Strategies}\label{sec:synthesis}
The goals, assumptions, and processes of matching, mediation analysis,
and principal stratification, when applied to implementation data from
an EdTech RCT, each differ from each other, but also overlap in
interesting ways.
In the case discussed here, the observational study and mediation
approaches appear to give the opposite conclusion from principal
stratification.
This section will compare and contrast the three methods in terms of
 their necessary identification assumptions, their model fitting
 mechanics, and interpretation of their results.

The fundamental difference between the observational study and
mediation approach, on the one hand, and principal stratification on
the other, is in the role of $H$, the intermediate variable.
An observational study or mediation analysis treats $H$ as a causal
agent, an intervention or exposure that affects $Y$.
In mediation analysis, it is also an (intermediate) outcome, affected
by $Z$.
The approach of principal stratification is quite nearly opposite:
$H$ is not an exposure, that ``happens to'' subjects, but rather its
potential value $H_T$ is a characteristic of the subjects.
Observed $H$ reveals what was there all along.

\subsection{Causal Identification}

Observational studies, and mediation analyses that build on them,
require ignorability assumptions such as Strong Ignorability or
Matched Ignorability.
These may be particularly problematic in our study, since students who
are more likely to request hints are also more likely to struggle with
the material, and hence score lower on the posttest.
We addressed this concern in three ways: first, we focused our attention
on the set of problems in the tutor in which students either requested
a hint or made an error, second, with propensity score matching, and
third with sensitivity analysis.
None of these approaches is bulletproof---they each require further
assumptions about how students request hints.
Eliminating problems in which students showed no signs of struggling
is likely to reduce the association between hint requests and
underlying academic ability, but not eliminate it.
For instance, stronger students may be better able to figure out a
difficult problem without access to hints or error feedback than
weaker students.
Matching depends both on the adequacy of the observed covariates to
capture baseline differences between $H=0$ and $H=1$ students, and the
sensitivity analyses we performed assumed that any un-measured
covariate would predict $H$ no better than pretest, or no better than
a covariate half as important as pretest.
These assumptions are impossible to verify from the data at hand;
instead, they require a keen understanding of student practice within
the tutor.

Mediation analysis builds on the matching design, and also requires an
ignorability assumption; it must
additionally overcome the hurdle of the almost complete lack of
overlap between hint request behavior in the control group (i.e. zero)
and in the treatment group.
Our approach of dichotomizing hint request behavior as $H$ provides a
way forward, at the cost of ignoring information that could be gleaned
from more continuous measures of hint requests.

In contrast, no ignorabilty assumptions are necessary for principal
stratification.
That said, an assumption similar to ignorability may play an important
role for interpretation.
Further, principal stratification estimators require a model for $Y(0)$
conditional on $\eta(1)$, which is unobserved.
On the other hand, model misspecification is, in principle,
detectable---an assumption of a particular family of models is not
``untestable.''
Still, the data will, in general, be unable to distinguish between certain
alternative models for $Y(0)$ conditional on $\eta(1)$, that may in
turn lead to qualitatively different conclusions.
In practice, analysts make the untestable assumption that $Y(0)$ conditional on $\eta(1)$ is
distributed similarly to $Y(1)$ conditional on $\eta(1)$.

An additional contrast of note between matching-based studies and
principal stratification relates to transparency.
However complex the process of devising a match may be, the resulting
design is simple and clear---subjects with $H=1$ are compared against
matches with $H=0$.
This transparency may help researchers assess the plausibility of the
match, and diagnose potential problems with resulting estimates.
Though PS does not depend on ignorability, it does depend on a highly
intricate and parametric modeling structure.
If any part of the model is sufficiently misspecified, the results may
be wrong.
Perhaps worse, \citet{feller2016principal} argued that even well-specified
models can give severely biased estimates.
Extensive model interrogation and checking is necessary for PS
analysis, but the complexity of the PS model makes this process
particularly difficult.
\citet{aoas} gives some examples of potentially fruitful model
checking procedures, which are also carried out in the online supplement.

\subsection{Results and Interpretation}

Our observational study found that hint requests have a negative effect: students who
requested hints more often than the \Sexpr{round((1-mixProp)*100)}th
percentile score lower on the posttest, after controlling for observed covariates---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
Mediation analysis essentially interprets this result in terms of the
overall CTA1 treatment effect.
Specifically, our analysis showed that CTA1 affected students'
posttest scores in (at least) two opposite ways: by allowing hints, it
lowered their posttest scores, but its other mechanisms increased
their scores by an even greater amount.

Like in the observational study and in mediation analysis, the
principal stratification analysis found a negative correlation
between hint requesting and posttest scores.
Unlike those two methods, PS understands $\eta(1)$ as a
student-specific baseline characteristic---albeit one that only
manifests in the treatment group.
Students who request more hints need more help, and thus score lower
on the posttest.
On the other hand, the PS analysis suggests that it's these students
who also receive more help---perhaps due to CTA1 hints, or other
features of the curriculum---and hence benefit more from CTA1.

@@

\subsection{Choosing Between Methods}

The previous discussion suggests three criteria for choosing between
the three approaches we've discussed here.
First is the question the researcher seeks to answer.
In their most straightforward interpretations, principal stratification
is a method for examining treatment effect heterogeneity, whereas
mediation analysis is a method for assessing causal mechanisms.
That said, PS may also shed light on potential causal mechanisms---if,
indeed, treatment effects are higher for students who request more
hints, then hints may play a role in the tutor's effectiveness.

A second criterion is the researcher's willingness to make
ignorability assumptions.
A researcher's confidence in her understanding of a data generating
process and the quality of observed covariates can translate into
confidence about the results of an observational study or mediation
analysis.
Conversely, researchers unwilling to consider untestable ignorability
assumptions will find PS an attractive alternative.

The final criterion is a researcher's comfort with complex statistical
models.
Researchers who are able and willing to experiment with a range of
models for a dataset, and perhaps devise tests of model fit tailored
to a specific problem, may be confident in the fit of a PS model.
Researchers who prefer transparency and non-parametric or
semi-parametric analysis may prefer matching studies.

Additional philosophical concerns come in to play as well.
For instance, unlike observational studies and PS, mediation analysis
depends on strictly counterfactual quantities, such as $\EE Y(0,H(1))$,
that can never occur.
Analogous quantities in PS, such as $\EE [Y(0)|H(1)]$ are
unobservable, but nevertheless refer to averages of potential outcomes
among actual experimental subjects.
These concerns are important but beyond the scope of the current
paper; for more discussion, see the citations in the introduction.

% \begin{table}
% \begin{tabular}{p{1in}p{1.5}p{2in}p{2in}}
% & Observational Study&Mediation Analysis &Principal Stratification\\
% \hline
% Research question & How does $H$ affect $Y$& What is the role of $Z$'s
%                                              effect on $H$ in its
%                                              effect on $Y$? & Does
%                                                               $Z$'s
%                                                               effect
%                                                               on $H$
%                                                               predict its
%                                                               effect
%                                                               on $Y$?\\

% \hline
% Role of $H$ & Causal & Causal & Predictive\\
% \hline
% Data used & treatment group & indirect effect: treatment group; direct
%                               effect: full dataset & full dataset\\
% \hline
% Untestable assumption &$H$ is unconfounded& $H$ is unconfounded & Distribution of $Y$
%                                               given $H_T$ in control
%                                               group\\
% \hline
% Other difficulties & Discrete $H$ necessary for mediation (in our
%                      framework) & Model fitting can be opaque and
%                                   erratic\\
% \end{tabular}
% \caption{Some contrasts between
% \end{table}

@@

A second challenge is extrapolation in mediation analysis.
in RCTs comparing computerized educational technology with traditional
pedagogy, members of the control group do not have computer log data
or access to the mechanisms that the log data measures.
This forms a serious barrier to mediation analysis.
However, by dichotomizing the mediator and matching, we were able to
re-define the causal question in such a way as to make it (possibly) answerable.

@@

\newpage
\appendix

\section{Rasch Mixture Model of Hint Usage}

To model hint requests with a mixture model, we modeled each hint
request as in \eqref{eq:rasch}.
Then, instead of modeling $\eta_i$ as normal, we assigned it a
two-component normal mixture model:
\begin{equation*}
\eta\sim
p_0\mathcal{N}(\mu_0,\sigma_0)+(1-p_0)\mathcal{N}(\mu_1,\sigma_1)
\end{equation*}
where we constrained $\mu_0<\mu_1$.

\subsection{Stan Code}
<<raschMixture,results='markup'>>=
  raschMix <- readLines('R/raschMix.stan')
  raschMix <- gsub('studEff','eta',raschMix)
  raschMix <- gsub('secEff','delta',raschMix)
cat(raschMix,sep="\n")
@
%\clearpage
\section{Covariate Balance Before and After Matching}
<<covariateBalanceTable,results='asis'>>=
library(RItools)
print(xtable(balMod,caption='Covariate balance (standardized differences and Z-scores) before and after the propensity score match.',label='tab:balance'),add.to.row=list(pos=list(-1),command='&\\multicolumn{3}{c}{Unmatched}&\\multicolumn{3}{c}{Matched}\\\\\n'))
@

\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{output/balancePlot.pdf}
\caption{Covariate balance (standardized differences) before and after
  the propensity score match.}
\label{fig:balance}
\end{figure}

%\clearpage
\section{Stan Code for PS Model}

%\clearpage
\subsection{Stan Code}
<<psCode,results='markup',cache=FALSE>>=
  pscode <- readLines('R/prinStratStan.stan')
  pscode <- gsub('studEff','eta',pscode)
  pscode <- gsub('secEff','delta',pscode)
pscode <- gsub('grad','hint',pscode)
cat(pscode,sep="\n")
@

%\clearpage
\section{Diagnostic Plots for Principal Stratification}

<<makeDiagnosticPlots,eval=FALSE>>=
source('R/diagnosticPlots.r')
@

\subsection{Posterior Predictive Density Plots}
\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{output/ppcYoverall.jpg}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{output/ppcYtrt.jpg}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{output/ppcYctl.jpg}
\end{figure}

%\clearpage
\subsection{Residual Plots}
\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{output/residPlotOverall.jpg}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{output/residPlotTrt.jpg}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{output/residPlotCtl.jpg}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{output/binnedplot.pdf}
\caption{Binned residual plot for model of hint requests.}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{output/qqPlots.pdf}
\caption{Q-Q plots for outcomes and fitted values}
\end{figure}
